### 2023-02-22 doc/response.tex

\revtext{The authors present a retrospective analysis of the COVID-19 epidemic at Princeton University and provide an extensive analysis that generally supports the hypothesis that a superspreading event led to a large uptick in case numbers after the introduction of the Omicron variant. Overall I like the paper. It's well written, and provide a detailed description of the application of a detailed computational model for policy support at an institutional level. It is therefore potentially a valuable contribution and I would support publication after some minor revisions.}

Thank you.

\revtext{Chief among these is the sensitivity of model results (i.e., calibration, and ultimate policy commentary) on the assumptions regarding test sensitivity. The authors claim "Our analysis highlights the power of mass asymptomatic testing" but the assumption is that a PCR test is 95\% sensitive after latency, and that case detection leads to full case isolation. These are both extremely strong assumptions, that are not conservative with respect to the claim. I recommend examining the results of Hellewell et al ("Estimating the effectiveness of routine asymptomatic PCR testing...") for strong evidence that sensitivity is much lower than claimed here. Also see Zachreson et al ("COVID-19 in low-tolerance border quarantine systems...") for a detailed individual-based implementation of dynamic test sensitivity.}

We agree that 95\% sensitivity seem too strong.
Hellewell et al estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
But this estimate implicitly accounts for latency, whereas we assume 95\% sensitivity during the infectious period after 0\% sensitivity during latency.
Therefore, while our sensitivity peaks at 92\% 5 days after infection, it decreases to 20\% by 10 days after infection (due to recovery from infections), which is considerably lower than what was estimated by Hellewell et al.
Therefore, while the individual-level sensitivity is not dynamic (a step function changing from 0 to 0.95 and back down to 0), the population-level sensitivity is dynamic.
We have added the following paragraph to our main text:

@@

We did this for computationally efficiency. We now explore R as low as 0.25.
We could explore even lower values of R but we feel that lower values may be unrealistic.

\revtext{4) Please explain "Thanksgiving" to non-Americans. This should be indicated visually in Fig. 2B}

We now explain Thanksgiving the first time it is mentioned:

``Thanksgiving---a national holiday in the US during which many students travel off campus''

@@

Thank you.

\revtext{I had two main concerns about the conclusions as presented that I hope the authors can address. The first is that while Princeton is an ideal ecosystem because of its asymptomatic screening protocols, the surrounding community is not necessarily one—this raises concerns about how differences in testing practices could be controlled for/accounted for in the correlation analysis.}

\revtext{Most estimates seem to a huge reduction in case ascertainment during the Omicron surge, which makes comparisons between the broader community and Princeton tricky for that time point. At the very least, I would encourage the authors to control for differences in denominator (e.g., converting to per 100K/day) and/or exploring test positivity as an additional metric to see if their observations still hold.}

We have revised our analysis and figures to account for changes in population sizes across the semester. See below for a detailed response.
We also present testing volumes over time on campus and show that they are reasonably stable across each semester.
We were not able to find information on testing at the county level and therefore were not able to explore test positivity as an additional metric.
However, we feel that test positivity can be even more sensitive to testing behavior in our cases.
For example, testing frequencies were doubled for undergraduate students during the initial Omicron break, but this change did not cause a doubling of case numbers, meaning that the ascertainment rate was already high. 
It seems more likely that the increased numbers of cases during this period is reflective of real outbreak dynamics, rather than testing behavior---on the other hand, positivity would suddenly decrease during this period due to changes in testing behavior.
Finally, strong cross correlations in case patterns across other counties in New Jersey and other large cities (New York City and Philadelphia) demonstrate the robustness of our observation.

### 2023-02-22 doc/university-covid.tex
am ready to resubmit this

Mathematical models have also been widely deployed in planning campus reopenings.
Researchers from various institutions in the US---including Cornell \citep{frazier2022modeling}, Emory \citep{lopman2020model}, Georgia Institute of Technology \citep{gibson2021surveillance}, and UC Berkeley \citep{brook2021optimizing}---modeled the feasibility of controlling the epidemic on their campuses and considered mass asymptomatic testing as their main intervention.
These modeling efforts helped identify key parameters for control, such as the testing turnaround time, and provided support for implementing similar measures at other institutions.
Coupling modeling efforts with real-life implementations in university campuses further provided unique opportunities to directly test model-based predictions of intervention effects in preventing the transmission of SARS-CoV-2 \citep{frazier2022modeling}---
each university campus offers a relatively well-controlled epidemic setting with a relatively homogeneously behaving population (especially among undergraduate students).
Campuses can also offer strong opportunities for control by non-pharmaceutical interventions, such as isolation and mask-wearing;
mass asymptomatic testing further provides robust ascertainment for epidemic sizes, allowing for accurate understanding of epidemic patterns.

@@

We begin with a descriptive analysis of the PU outbreak (\fref{princeton}), then present modeling analyses of the individual epidemics during 2020--2022.
PU is located in Mercer County, New Jersey, USA;
the population comprises 5267 undergraduate students, 2946 graduate students, and around 7000 faculty and staff members.
For simplicity, we divided the epidemic into four time periods representing four semesters across two academic years: Fall 2020--2021 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020--2021 (January 16, 2021--May 14, 2021; \fref{princeton}B), Fall 2021--2022 (August 14, 2021--December 31, 2021; \fref{princeton}C), and Spring 2021--2022 (January 1, 2022--March 18, 2022; \fref{princeton}D).
Throughout the majority of the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
Those who tested positive were required to isolate for at least 10 days after symptom onset or test date (whichever was longer) and were released when they had been at least 48 hours with improving or resolving symptoms as per New Jersey Department of Health guidance.
PCR positives were exempt from asymptomatic testing for 90 days.
Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine status is up-to-date. This in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
In Supplementary Figure S1, we show the testing volumes over time.
Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to either quarantine or test more frequently according to the close contacts' vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
Changes in testing frequency and other intervention measures throughout the study period reflected various factors, including the impact of COVID-19 cases on continuity of operations or continuity of teaching; on severity of disease on campus; the capacity of testing and the healthcare system; and hospitalization rates on campus and in the area.
All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.

@@

During the \textbf{fall semester of the 2020--2021 academic year}, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
All classes were held virtually, and so only a few undergraduate students remained on campus ($<300$).  
Both undergraduate and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
A sudden decrease in the number of cases around Thanksgiving---a national holiday in the US during which many students---partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S2A)---we exclude the undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.

In the beginning of the \textbf{spring semester of the 2020--2021 academic year}, the number of cases suddenly increased before classes started (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
Returning students were required to be tested and quarantine for 7 days regardless of their returning location.
Most classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
Some smaller classes were held in-person, but required social distancing (thereby limiting the size of the class) and mask-wearing at all times.
The number of cases persisted at similar levels to the fall semester and eventually decreased as classes ended and students went home---the decrease in the number of cases in PU also coincided with the decrease in the number of cases in Mercer County.
The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).
This ordering is robust to differences in population sizes: 37 cases per 1000 among faculty and staff members, 34.3 cases per 1000 among undergraduate students, and 14.5 cases per 1000 among graduate students (Supplementary Figure S2B).

For the \textbf{fall semester of the 2021--2022 academic year}, all students and faculty and staff members were required to be vaccinated, with very few medical and religious exemptions.
By the beginning of the semesters, 97\% of undergraduate students, 96\% of graduate students, and 94\% of faculty and staff members were vaccinated.
Vaccinees were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
In-person classes and social events fully resumed on campus, though all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
The number of cases remained similar to previous semesters until November when cases began to increase, primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
In order to prevent transmission, testing frequency was increased to twice a week for undergraduate students on November 27th, 2021; the size of non-academic gatherings were also limited to 20 people.
The number of cases decreased slightly as classes ended but soon increased again as the Omicron (BA.1) variant began to spread on campus and in Mercer county.
The total number of reported cases per 1000 remained high for all three population groups: 74 cases per 1000 among faculty and staff members, 63.5 cases per 1000 among graduate students, and 52.4 cases per 1000 among undergraduate students (Supplementary Figure S2C).

For the \textbf{spring semester of the 2021--2022 academic year}, all eligible students and faculty and staff members were required to have obtained booster vaccination.
By the beginning of the semesters, 65\% of undergraduate students, 71\% of graduate students, and 82\% of faculty and staff members were boosted.
Undergraduate students were still required to be tested twice a week to prevent the additional spread of the Omicron variant.
The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events were no longer limited to 20 people;
in addition, the testing frequency was reduced to once a week.
Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).
The timing of this outbreak also coincided with a rapid increases in the Omicron subvariant BA.2 cases---the proportion of BA.2 subvariant reached 93.5\% (372/398) compared to 26.9\% (14/52) from the previous week.
On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
Cases were largely concentrated among undergraduate students during this semester: 309 cases per 1000 among undergraduate students, 106 cases per 1000 among faculty and staff members, and 65.3 cases per 1000 among graduate students.

@@

Infection processes are modeled based on standard compartmental structures (Supplementary Figure S8).
Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
Exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections for $D_p = 3$ days on average.
Presymptomatic individuals can then either remain asymptomatic with probability $p_a = 0.4$ or develop symptoms with the remaining probability of $1-p_a = 0.6$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s=3$) and equal transmission rates.
Recovered individuals are assumed to be immune to reinfections throughout a semester.
Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments to allow for more realistic and narrower distributions than the exponential distribution \citep{brett2020transmission}.
Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}:
more specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
Assumed parameters are broadly consistent with other models of SARS-CoV-2 \citep{brett2020transmission,lavezzo2020suppression}.

@@

The 95\% sensitivity assumption may seem too high. 
For example, \cite{hellewell2021estimating} estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
These estimates are considerably lower than our assumption because their estimates implicitly account for the latent period.
At the individual level, we assume that an infected individual has no detectable infection (0\% sensitivity) during their latent period and 95\% sensitivity during their various stages of infectious periods.
At the population level, this assumption translates to a peak sensitivity of 92\% by 4 days after infection, decreasing to 20\% sensitivity by 10 days after infection.
Our assumption leads to a much lower PCR sensitivity 10 days after infection because we only model the PCR sensitivity during the infectious period.
In reality, PCR can detect infections even after a person stops shedding infectious virus---we did not include this component in our model because it would not affect the effectiveness of isolation strategy in reducing transmission.

@@

In this study, we use this model to retrospectively analyze past outbreaks.
First, we try to match our model to epidemic patterns seen on campus for the first three semesters, during which there was limited campus transmission, by varying the contact reproduction number $\Rc$ and the amount of community contact $\theta$ and holding all other parameters constant.
For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume, for simplicity, that all other parameters are equal across different groups 
We further assume that the population mixes homogeneously.
While these assumptions are most parsimonious, epidemiological parameters and mixing patterns likely differ across groups (e.g., undergraduate students are more likely to infect undergraduate students and also remain asymptomatic).
Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.

@@

For \textbf{spring 2020--2021}, we simulate the model assuming 8000 individuals (3000 undergraduate students, 2000 graduate students, and 3000 faculty and staff members) on campus with 5000 of them participating in asymptomatic testing twice a week.
We further assume that 4000 individuals (3000 undergraduate students and 1000 graduate students) returned to campus over 14 days (January 16, 2021--January 29, 2021).
In the beginning of the semester, all returning students were required to quarantine in their rooms for 14 and tested upon returning by the university---in our model, this was implemented by preventing returning students from getting infected or infecting other individuals.
Finally, to match the initial influx of cases, we assume that 1\% of both returning and on-campus populations are infected at the beginning of simulation (January 16, 2021).

@@

Epidemiological conditions and intervention measures changed throughout the spring semester of the 2021--2022 academic year.
We therefore extend the model to account for these alterations and focus on the outbreak patterns among undergraduate students.
First, based on \citep{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the early Omicron variant by 10\% and 70\%, respectively. 
We also assume that the transmissibility of Omicron is reduced proportionally following the previously assumed 90--to--20 ratio for the Delta variant;
in other words, two and three doses of vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
The immunity from the third dose is assumed to take 7 days to develop \citep{moreira2022safety} and wane at the same rate as before (in this case, 70\% to 39\% in 20 weeks).
Finally, the isolation period is reduced to 5 days.

@@

Despite the simplicity of the analysis, our study provides important lessons for controlling SARS-CoV-2 and similar outbreaks on university campuses in general.
First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning---even if PCR testing may have lower sensitivity than what we assumed here \citep{hellewell2021estimating}, mass asymptomatic testing can still help track ongoing epidemic dynamics in real time.
Combining other interventions measures, such as social distancing, mask wearing, and vaccination, can help provide a safe means of reopening university campuses---but the extent to which these interventions are implemented will necessarily depend on resource availability.
Second, we expect immune waning and superspreading to continue to play important roles in driving campus transmission---keeping vaccine statuses up-to-date within the campus community will be critical moving forward.
Third, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both potentially drive transmission in each other---however, the degree to which infections spread from campus to community remains uncertain.
Finally, intervention measures placed on campuses must continue to adapt and change to reflect changes in epidemiological conditions.

The emergence of new variants---in particular, their ability to evade prior immunity and transmit better---continues to add uncertainty to the future controllability of the ongoing SARS-CoV-2 pandemic.
Nonetheless, as population-level immunity increases (either due to infection or vaccination), we are (hopefully) transitioning to an endemic phase, during which COVID-19 is expected to become less severe \citep{lavine2021immunological}.
Many institutions have so far sought to minimize infections on their campuses early in the pandemic, thereby implementing as many interventions as possible---but such measures can be difficult to maintain for prolonged periods both economically and societally.
As the Omicron variant began to spread, many campuses---including the PU campus---opted in for less intense interventions, reflecting difficulties in controlling the spread and a lack of severe cases among the majority of vaccinated students.
As we continue to transition to future phases of the pandemic, the expectations for reopening campuses (e.g., whether to minimize infections on campuses) need to be re-evaluated, accounting not only for changes in epidemic dynamics but also for our perception of the pathogen.
The answers to these questions ultimately depend on the landscape of SARS-CoV-2 immunity and its future evolutionary dynamics \citep{saad2020immune,baker2021limits}.

>>>>

Mathematical models have also been widely deployed in planning campus reopenings.
Researchers from various institutions in the US---including Cornell \citep{frazier2022modeling}, Emory \citep{lopman2020model}, Georgia Institute of Technology \citep{gibson2021surveillance}, and UC Berkeley \citep{brook2021optimizing}---modeled the feasibility of controlling the epidemic on their campuses and considered mass asymptomatic testing as their main intervention.
Mathematical models also played crucial roles in helping to make decisions about reopening, ensuring that infrastructure (e.g., isolation methods and spaces, supporting isolation and quarantine with food, etc.) and levels of mitigation (masks, ventilation, testing frequency, etc.) were adequate and appropriate.
These modeling efforts helped identify key parameters for control, such as the testing turnaround time, and provided support for implementing similar measures at other institutions.
Coupling modeling efforts with real-life implementations in university campuses further provided unique opportunities to directly test model-based predictions of intervention effects in preventing the transmission of SARS-CoV-2 \citep{frazier2022modeling}---
each university campus offers a relatively well-controlled epidemic setting with a relatively homogeneously behaving population (especially among undergraduate students).
Campuses can also offer strong opportunities for control by non-pharmaceutical interventions, such as isolation and mask-wearing;
mass asymptomatic testing further provides robust ascertainment for epidemic sizes, allowing for accurate understanding of epidemic patterns.

@@

We begin with a descriptive analysis of the PU outbreak (\fref{princeton}), then present modeling analyses of the individual epidemics during 2020--2022.
PU is located in Mercer County, New Jersey, USA;
the population comprises 5267 undergraduate students, 2946 graduate students, and around 7000 faculty and staff members.
For simplicity, we divided the epidemic into four time periods representing four semesters across two academic years: Fall 2020--2021 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020--2021 (January 16, 2021--May 14, 2021; \fref{princeton}B), Fall 2021--2022 (August 14, 2021--December 31, 2021; \fref{princeton}C), and Spring 2021--2022 (January 1, 2022--March 18, 2022; \fref{princeton}D).
Throughout the majority of the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
Those who tested positive were required to isolate for at least 10 days after symptom onset or test date (whichever was longer) and were released when they had been at least 48 hours with improving or resolving symptoms as per New Jersey Department of Health guidance;
the isolation duration switched to 5 days on January 14, 2022, for all vaccinated people.
PCR positives were exempt from asymptomatic testing for 90 days.
Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine status is up-to-date. This in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
In Supplementary Figure S1, we show the testing volumes over time---the testing volumes remained roughly constant within each semester, except when testing frequency changed, which we note in the paper.
However, changes in testing frequency were not associated with sudden changes in cases, meaning that the patterns in case trajectories likely reflect patterns of spread, rather than testing behavior.
Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to test more frequently and quarantine as applicable, according to the close contacts’ vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
Changes in testing frequency and other intervention measures throughout the study period reflected various factors, including the impact of COVID-19 cases on continuity of operations or continuity of teaching; on severity of disease on campus; the capacity of testing and the healthcare system; and hospitalization rates on campus and in the area.
All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.

@@

During the \textbf{fall semester of the 2020--2021 academic year}, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
All classes were held virtually, and so only a few undergraduate students remained on campus ($<300$).  
Both undergraduate and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
A sudden decrease in the number of cases around Thanksgiving---a national holiday in the US during which many students travel off campus---partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S2A)---we exclude the undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.

In the beginning of the \textbf{spring semester of the 2020--2021 academic year}, the number of cases suddenly increased before classes started (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
Returning students were required to be tested upon arrival and quarantine for 7 days regardless of their returning location.
Most classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
Some smaller classes were held in-person, but required social distancing (thereby limiting the size of the class) and mask-wearing at all times.
The number of cases persisted at similar levels to the fall semester and eventually decreased as classes ended and students went home---the decrease in the number of cases in PU also coincided with the decrease in the number of cases in Mercer County.
The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).
This ordering is robust to differences in population sizes: 37 cases per 1000 among faculty and staff members, 34.3 cases per 1000 among undergraduate students, and 14.5 cases per 1000 among graduate students (Supplementary Figure S2B).

For the \textbf{fall semester of the 2021--2022 academic year}, all students and faculty and staff members were required to be vaccinated with the COVID vaccine primary series, with very few medical and religious exemptions.
By the beginning of the semesters, 97\% of undergraduate students, 96\% of graduate students, and 94\% of faculty and staff members were vaccinated.
Vaccinees were required to be tested once a week, while unvaccinated and partially vaccinated individuals were required to be tested twice a week.
In-person classes and social events fully resumed on campus, though all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching any size of class if social distancing can be maintained at all times).
The number of cases remained similar to previous semesters until November when cases began to increase, primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
In order to prevent transmission, testing frequency was increased to twice a week for undergraduate students on November 27th, 2021; the size of non-academic gatherings were also limited to 20 people.
The number of cases decreased slightly as classes ended but soon increased again as the Omicron (BA.1) variant began to spread on campus and in Mercer county.
The total number of reported cases per 1000 remained high for all three population groups: 74 cases per 1000 among faculty and staff members, 63.5 cases per 1000 among graduate students, and 52.4 cases per 1000 among undergraduate students (Supplementary Figure S2C).

For the \textbf{spring semester of the 2021--2022 academic year}, all eligible students and faculty and staff members were required to have obtained booster vaccination.
By the beginning of the semesters, 65\% of undergraduate students, 71\% of graduate students, and 82\% of faculty and staff members were boosted.
Undergraduate students were still required to be tested twice a week to prevent the additional spread of the Omicron variant.
The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events and event sizes were no longer limited to 20 people;
in addition, the testing frequency was reduced to once a week (but not for unvaccinated individuals).
Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).
The timing of this outbreak also coincided with a rapid increases in the Omicron subvariant BA.2 cases---the proportion of BA.2 subvariant reached 93.5\% (372/398) compared to 26.9\% (14/52) from the previous week.
On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
Cases were largely concentrated among undergraduate students during this semester: 309 cases per 1000 among undergraduate students, 106 cases per 1000 among faculty and staff members, and 65.3 cases per 1000 among graduate students.

@@

Infection processes are modeled based on standard compartmental structures (Supplementary Figure S8).
Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
Exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections for $D_p = 3$ days on average.
Presymptomatic individuals can then either remain asymptomatic with probability $p_a = 0.4$ or develop symptoms with the remaining probability of $1-p_a = 0.6$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s=3$) and equal transmission rates.
Recovered individuals are assumed to be immune to reinfections throughout a semester.
Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments.
Dividing each infection stage into subcompartments allows for the duration of infection to have narrower (and more realistic) distributions than the exponential distribution \citep{brett2020transmission}.
Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}:
more specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
Assumed parameters are broadly consistent with other models of SARS-CoV-2 \citep{brett2020transmission,lavezzo2020suppression}.

@@

The 95\% sensitivity assumption may seem too high. 
For example, \cite{hellewell2021estimating} estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
These estimates are considerably lower than our assumption because their estimates implicitly account for the latent period.
At the individual level, we assume that an infected individual has no detectable infection (0\% sensitivity) during their latent period and 95\% sensitivity during their various stages of infectious periods.
At the population level, this assumption translates to a peak sensitivity of 92\% by 4 days after infection, decreasing to 20\% sensitivity by 10 days after infection.
Our assumption leads to a much lower PCR sensitivity 10 days after infection because we only model the PCR sensitivity during the infectious period.
In reality, PCR can also detect viral nucleic acids even after a person stops shedding infectious virus, but since these nucleic acids cannot contribute to person-to-person transmission or affect the effectiveness of the isolation strategy, we did not include this component in our model.

@@

In this study, we use this model to retrospectively analyze past outbreaks.
First, we try to match our model to epidemic patterns seen on campus for the first three semesters, during which there was limited campus transmission, by varying the contact reproduction number $\Rc$ and the amount of community contact $\theta$ and holding all other parameters constant.
For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume, for simplicity, that all other parameters are equal across different groups.
We further assume that the population mixes homogeneously.
While these assumptions are most parsimonious, epidemiological parameters and mixing patterns likely differ across groups (e.g., undergraduate students are more likely to infect undergraduate students and also remain asymptomatic).
Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.

@@

For \textbf{spring 2020--2021}, we simulate the model assuming 8000 individuals (3000 undergraduate students, 2000 graduate students, and 3000 faculty and staff members) on campus with 5000 of them participating in asymptomatic testing twice a week.
We further assume that 4000 individuals (3000 undergraduate students and 1000 graduate students) returned to campus over 14 days (January 16, 2021--January 29, 2021).
In the beginning of the semester, all returning students were required to quarantine in their rooms for 14 days and tested upon returning by the university---in our model, this was implemented by preventing returning students from getting infected or infecting other individuals.
Finally, to match the initial influx of cases, we assume that 1\% of both returning and on-campus populations are infected at the beginning of simulation (January 16, 2021).

@@

Epidemiological conditions and intervention measures changed throughout the spring semester of the 2021--2022 academic year.
We therefore extend the model to account for these alterations and focus on the outbreak patterns among undergraduate students.
First, based on \cite{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the early Omicron variant by 10\% and 70\%, respectively. 
We also assume that the transmissibility of Omicron is reduced proportionally following the previously assumed 90--to--20 ratio for the Delta variant;
in other words, two and three doses of vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
The immunity from the third dose is assumed to take 7 days to develop in our model \citep{moreira2022safety} and wane at the same rate as before (in this case, 70\% to 39\% in 20 weeks).
Finally, the isolation period was reduced to 5 days---the actual change was implemented on January 14, 2022 but we keep the 5 day isolation period throughout our simulations, which begins on January 7, 2022, for simplicity.

@@

Despite the simplicity of the analysis, our study provides important lessons for controlling SARS-CoV-2 and similar outbreaks on university campuses in general.
First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning---even if PCR testing may have lower sensitivity than what we assumed here \citep{hellewell2021estimating}, mass asymptomatic testing can still help track ongoing epidemic dynamics in real time.
Combining other interventions measures, such as social distancing, mask wearing, and vaccination, can help provide some measures to consider for restoring operations on campuses---but the extent to which these interventions are implemented will necessarily depend on resource availability.
Second, we expect immunity waning and superspreading to continue to play important roles in driving campus transmission---keeping vaccine statuses up-to-date within the campus community will be critical moving forward.
Third, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both potentially drive transmission in each other---however, the degree to which infections spread from campus to community remains uncertain.
Finally, intervention measures placed on campuses must continue to adapt and change to reflect changes in epidemiological conditions.
We note that the generality of our conclusions will necessarily depend on specific campus settings.

The emergence of new variants---in particular, their ability to evade prior immunity and transmit better---continues to add uncertainty to the future controllability of the ongoing SARS-CoV-2 pandemic.
Nonetheless, as population-level immunity increases (either due to infection or vaccination), we are transitioning to an endemic phase, during which COVID-19 is expected to become less severe \citep{lavine2021immunological}.
Understanding the the landscape of SARS-CoV-2 immunity and its impact on evolutionary dynamics will be critical to predicting future outbreak dynamics \citep{saad2020immune,baker2021limits}.
